Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-26 survey/survey.tex
intro

@@

{\begin{quote} Mathematical models of computation are abstract
  constructions, by their nature unfettered by physical laws. However,
  if these models are to give indications that are relevant to
  concrete computing, they must somehow capture, albeit in a selective
  and stylized way, certain general physical restrictions to which all
  concrete computing processes are subjected~\cite{Toffoli:1980}.
\end{quote}}

@@

The technical development derives from ideas which are implicit in the
works of \cite{Toffoli:1980, Zuliani:2001:LR,
  malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.
(1) Functions whose output entropy is equal to their input entropy are
information preserving functions.  (2) A function
$f : b_1 \rightarrow b_2$ is \emph{logically reversible} if there
exists an inverse function $f^{\dagger}$ such that for any input
$v_1:b_1$ if there exists $v_2:b_2$ such that $f(v_1)=v_2$ \emph{iff}
$f^{\dagger}(v_2)=v_1$.  (3) Logically reversible functions are
information preserving.

>>>>

What kind of operations can computers perform? This question was
answered several times in the last hundred years, where each answer
proposes an abstract \emph{model of computation} that specifies
allowable operations and (usually) their cost. The emerging consensus,
reflected in both early models of computations such as the Turing
Machine and the RAM as well as in the early Von Neumann and in modern
computer architectures, is that basic computer operations include
logical operations like conjunction, disjunction, and negation, as
well as reading from and writing to a large (infinite) collection of
memory locations.

No doubt, this consensus has been successful. And no doubt, these
operations \emph{can} indeed be performed on a computer. Yet, today,
with a possible quantum computing revolution in sight and with
unprecedented explosion in embedded computers and cyber-physical
systems, there are reasons to re-think this foundational question
again. In fact, the calls to re-think this foundational question have
been proclaimed by physicists for more than thirty years as the
following two quotes testify:

\begin{quote}
  Mathematical models of computation are abstract constructions, by
  their nature unfettered by physical laws. However, if these models
  are to give indications that are relevant to concrete computing,
  they must somehow capture, albeit in a selective and stylized way,
  certain general physical restrictions to which all concrete
  computing processes are subjected~\cite{toffoli:1980}.

  Another thing that has been suggested early was that natural laws
  are reversible, but that computer rules are not. But this turned out
  to be false; the computer rules can be reversible, and it has been a
  very, very useful thing to notice and to discover that. This is a
  place where the relationship of physics and computation has turned
  itself the other way and told us something about the possibilities
  of computation. So this is an interesting subject because it tells
  us something about computer rules
  \ldots~\cite{springerlink:10.1007/bf02650179}.
\end{quote}

These quotes by Toffoli and Feynman both highlight two obvious
observations: (i) all the operations that a computer performs reduce
to basic physical operations; and (ii) there is a mismtach between the
logical operations of a typical model of computation (which are
logically irreversible) and the fundamental laws of physics (which are
reversible). One could certainly dismiss the mismatch but our thesis
is that there is great value in revisiting models of computation to be
in harmony with the laws of physics.

\begin{comment}
\begin{quote}
  Ed Fredkin pursued the idea that information must be finite in
  density. One day, he announced that things must be even more simple
  than that. He said that he was going to assume that information
  itself is conserved. “You’re out of you mind, Ed.” I
  pronounced. “That’s completely ridiculous. Nothing could happen in
  such a world. There couldn’t even be logical gates. No decisions
  could ever be made.” But when Fredkin gets one of his ideas, he’s
  quite immune to objections like that; indeed, they fuel him with
  energy. Soon he went on to assume that information processing must
  also be reversible — and invented what’s now called the Fredkin
  gate. (Minsky 1999)

  In other terms, what is so good in logic that quantum physics should
  obey?  Can't we imagine that our conceptions about logic are wrong,
  so wrong that they are unable to cope with the quantum miracle?
  [\ldots] Instead of teaching logic to nature, it is more reasonable
  to learn from her. Instead of interpreting quantum into logic, we
  shall interpret logic into quantum (Girard 2007).
\end{quote}
\end{comment}

@@

{\begin{quote} Mathematical models of computation are abstract
  constructions, by their nature unfettered by physical laws. However,
  if these models are to give indications that are relevant to
  concrete computing, they must somehow capture, albeit in a selective
  and stylized way, certain general physical restrictions to which all
  concrete computing processes are subjected~\cite{toffoli:1980}.
\end{quote}}

@@

The technical development derives from ideas which are implicit in the
works of \cite{toffoli:1980, Zuliani:2001:LR,
  malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.
(1) Functions whose output entropy is equal to their input entropy are
information preserving functions.  (2) A function
$f : b_1 \rightarrow b_2$ is \emph{logically reversible} if there
exists an inverse function $f^{\dagger}$ such that for any input
$v_1:b_1$ if there exists $v_2:b_2$ such that $f(v_1)=v_2$ \emph{iff}
$f^{\dagger}(v_2)=v_1$.  (3) Logically reversible functions are
information preserving.
